Solution of an Acoustic Transmission Inverse Problem by Extended Inversion: Theory

### 2021-10-29 repro/papers/discreptheory/paper.tex
arXiv

\title{Solution of an Acoustic Transmission Inverse Problem by Extended Inversion}
\author{William W. Symes\footnotemark[1], Huiyi Chen\footnotemark[2],
  and Susan E. Minkoff\footnotemark[2]}
\address{\footnotemark[1]PO Box 43, Orcas, WA 98280 USA\\
  \footnotemark[2]Department of Mathematical Sciences, FO 35, 
  University of Texas at Dallas, Richardson, TX 75080 USA}

@@

\righthead{Discrepancy}

\maketitle
\begin{abstract}
  Study of an extremely simple single-trace transmission example shows
  how an extended source formulation of full waveform inversion can
  produce an optimization problem without spurious local minima
  (``cycle skipping''), hence efficiently solvable via Newton-like
  local optimization methods. The data consist of a single trace
  extracted from a causal pressure field, propagating in a fluid
  according to linear acoustics and recorded at a given distance from
  a transient point energy source. The velocity or slowness of the
  fluid is presumed homogeneous, and the target source intensity
  (``wavelet'') is presumed quasi-impulsive, with zero energy for time
  lags greater than a specified maximum lag (specified support). The
  data are assumed noisy, with known RMS signal-to-noise ratio. The
  inverse problem is: from the recorded trace, recover both the sound
  velocity and source wavelet with specified support, so that the data
  is fit with prescribed RMS relative error. After showing explicitly that
  the obvious least squares objective has multiple large residual
  minimizers, we introduce an extended approach. The source is extended by
  permitting energy to spread in time. The spread is controlled by
  adding a weighted mean square of the extended source wavelet to the
  data misfit, to produce the extended inversion objective. We choose
  a weight operator to penalize source energy away from zero lag via
  multiplication by the time lag. The resulting objective function and
  its gradient can be computed explicitly. We show that the error in 
  slowness component of any local minimizer is bounded by multiple of
  the maximum lag and the assumed noise level. In particular, there
  are no spurious local minima. We also show that the projection of
  the extended wavelet onto the interval of support yields a solution
  of the inverse problem as posed above, with appropriate choice of
  maxiumum lag and RMS relative error. In several appendices, we
  describe a constructive algorithm for controlling the penalty
  weight, the relation between maxiumum lag and RMS central frequency
  via the Heisenberg principle, and some structural details which the
  very simple inverse problem discussed here shares with more
  prototypical problems in industrial and academic seismology.

  %which may be interpreted as a wavelength.
  %The derivation shows
  %several important features of all similar extended source
  %algorithms. For example, nested optimization, with the source
  %estimation in the inner optimization (variable projection method),
  %is essential. The choice of the penalty operator, controlling the
  %extended source degrees of freedom, is critical: in order to produce
  %an objective immune from cycle-skipping, the penalty operator must
  %be (pseudo-)differential. The penalty weight is chosen dynamically,
  %according to a version of the Discrepancy Principle, in order to
  %accelerate convergence to a near-global minimum. Like many other
  %extended formulations, the example problem considered here permits
  %the discrepancy criterion to be satisfied with zero penalty
 % weight. This unusual feature removes a common obstacle to the use of
  %the discrepancy principle, namely the determination of a
  %satisfactory initial estimate of the penalty weight: the initial
 % weight may be set to zero, and the dynamic algorithm updates the
 % weight until convergence is achieved.

@@

\section{Introduction}
Model-based parameter estimation by least squares data fitting is a
widely used and successful technique for data analysis in science and
engineering \cite{Bertero:98,Vogel:02}, particularly in the
geosciences \cite{Parker:94,Tarantola:05}. In its application to
seismology, least-squares data fitting has come to be known as Full
Waveform Inversion (FWI), and is now well-established as a useful tool
for probing the earth's subsurface
\cite[]{VirieuxOperto:09,Fichtner:10,Schuster:17}. However, in its
application to seismology, FWI encounters a serious
practical obstacle, so-called ``cycle-skipping''. This term signifies
the tendency of the mean-square error function to exhibit many
stationary points far from the global minimizer (we
show an explicit example of this phenomenon later in this paper). Because of the
typical dimensions of field FWI problems, only local gradient-based
minimization algorithms are computationally practical. These find
stationary points (or local minima), so may stagnate at suboptimal and geologically
uninformative earth models. Local descent methods avoid suboptimal
stagnation only if initial models are already quite close to optimal,
in the sense of predicting the arrival times of seismic events to
within a small multiple of the dominant data wavelength
\cite[]{GauTarVir:86,VirieuxOperto:09,Plessix:10}.

This paper examines one of a number of proposed alternatives to least
squares FWI, based on an extension of the model-to-data mapping
(``extended inversion''), in the context of 
a very simple acoustic transmission inverse problem: the sound
velocity in a fluid is to be determined from a single recording of a
pressure wave propagated over a known distance, originating at a known
location around a known time. We show precisely
how descent-based FWI fails to solve this problem absent a very good
initial model estimate, and precisely why extended inversion succeeds, using the same type
of optimization, and from essentially arbitrary initial model
estimate. We give rigorous estimates for the accuracy of the sound
velocity determination in terms of the time extent of the energy
excitation and the signal-to-noise ratio of the data. We also illustrate
these mathematical conclusions with numerical
examples.

The single-trace transmission problem studied in this paper is a
drastically simplified cartoon of actual seismic prospecting. It seems
worth careful study nonetheless for two reasons: it exhibits the
phenomenon of cycle-skipping, and its simplicity allows the properties
of both FWI and extended inversion approaches to be established
rigorously. The underlying mathematical structures are common to many
inverse problems in wave propagation in which wave velocities are
amongst the unknowns. For example, we offer an ad hoc calculation of
the gradient of the extended inversion objective, taking advantage of
the simplicity of the problem context. However we also show, in an
appendix, how that calculation hides an abstract structure shared by
many other problems.% - see for example
%\cite{Symes:IPTA14,tenKroode:IPTA14,Symes:EAGE15,HuangSymes:SEG15a}.

In the next (``Overview'') section, we describe the conceptual
framework of this inverse problem, review the related literature, and
state the essence of our main results, leaving out a few technical
details that are supplied later. Numerical illustrations of these
results occupy the following (``Examples'') section. The fourth
(``Findings'') section of the paper contains precise statements of the
mathematical results, derivations, and supporting discussion. Three
appendices describe the algorithm used to control the penalty weight
in our formulation of extended inversion, further refinement of error
estimates based ond the Heisenberg inequality, and the abstract
structure of extended inversion implicit in the derivations presented
in the ``Findings'' section.

\section{Overview}

The physical setting for our discussion is the propagation of
small-amplitude pressure waves through a
homogeneous acoustic fluid occupying $\bR^3$ \cite[]{Frie:58}. The
fluid's salient characteristic for present purposes is the
slowness, or reciprocal wave velocity, $m > 0$. A source of acoustic
energy localized at a (``source'') point $\bx_s \in \bR^3$ radiates uniformly in all
directions with time-dependent intensity $w(t)$, generating a pressure
field $p(\bx,t)$. Both the 
pressure field $p$ and the intensity (``wavelet'') $w$ vanish at large
negative time. A receiver records the pressure values at a point
$\bx_r \in \bR^3$, a distance
$r>0$ from the source position $\bx_s$, between the times $t_{\rm min}$ and
$t_{\rm max}$. According to linear acoustics, the pressure field
solves the the wave equation \cite[]{Frie:58}:
\begin{eqnarray}
  \label{eqn:awe}
  \left(m^2\frac{\partial^2 p}{\partial t^2} - \nabla^2\right) p(\bx,t) &=&
                                                                         w(t)\delta(\bx-\bx_s) \nonumber\\
  p(\bx,t)&=&0, t\ll 0.
\end{eqnarray}
As the system \ref{eqn:awe} is linear and autonomous, the solution is
the convolution of the source
wavelet $w$ with the Green's function or fundamental solution. For
constant $m$, the result is the well-known expression
(\cite{CourHil:62}, Chapter VI, section 12, equation 47):
\begin{equation}
  \label{eqn:homsol}
  p(\bx,t) = \frac{1}{4\pi |\bx-\bx_s|}w\left(t-m|\bx-\bx_s|\right),
\end{equation}
so the predicted pressure at the receiver point $\bx_r$ is
\begin{equation}
\label{eqn:mod}
p(\bx_r,t) = \frac{1}{4\pi r}w\left(t-mr\right)  = F[m]w(t)
\end{equation}
The mapping $F[m]$ so defined is (ignoring the amplitude factor
$1/(4\pi r)$), an $m$-dependent time shift. This time shift operator is the basis of many illustrations of
the cycle-skipping phenomenon (for example, \cite{VirieuxOperto:09},
Figure 7), so it is unsurprising that an analysis of cycle-skipping
can be based on the simple modeling operator described above.

If the wavelet $w$ is square-integrable, so is the predicted
recorded pressure $F[m]w$ - its $L^2$ norm is the cumulative power trasferred
from the fluid to the receiver \cite[]{SantosaSymes:00}. Thus this relation defines a map
$F: \bR^+ \times L^2(\bR) \rightarrow
L^2([t_{\rm min},t_{\rm max}])$, linear in its second argument.

A naive version of the inverse problem investigated in this paper
would be: given recorded data
$d \in L^2([t_{\rm min},t_{\rm max}])$, find $m \in \bR^+$ and $w \in L^2(\bR)$ so that
$F[m]w \approx d$. Clearly this problem statement is not interesting as it
stands, since $F[m]$ is surjective for every $m \in \bR^+$: fitting
the data does not constrain the slowness $m$ at all.

To be useful, the problem statement must be augmented with a
constraint on the model $(m,w)$.  One natural constraint is to assume
that the support of $w$ is limited: choose $\lambda > 0$ and search
for $w$ with $\mbox{supp }w \subset [-\lambda,\lambda]$. The quantity
$\lambda$ is to be considered ``small''. Under some circumstances, it
is related to a mean wavelength of $w$ (Appendix B).

An assumption of small support may be justified as
follows. In practical seismic data processing, actual sources may act
over considerable time intervals. An example from exploration
seismology is the airgun, the most commonly used
marine source: an airgun (or array of airguns, the usual apparatus)
produces an oscillating pressure pulse that dies away
slowly. It is commonplace to estimate this pulse, usually by recording
the emitted pressure field at a position where it should nominally be isolated from other
signal, then deconvolve it
from the data by safeguarded Fourier division or other means. This
so-called signature deconvolution results in modified data
corresponding to a source equal to signature deconvolution of the
pulse estimate itself. The deconvolved source pulse approximates an
impulse ($\delta(t)$): it cannot have point support, due to the finite
frequency nature of seismic data and the Heisenberg principle (see
Appendix B), but its amplitudes are typically small outside of a small
much smaller time interval than is the case for the original pulse
estimate. Thus after a small modification, the deconvolved source
pulse has small support.

For a contemporary description of view on signature deconvolution as
practiced in the seismic industry, see \cite{SheriffGeldart:1995},
section 9.5 and \cite{Yil:01}, Chapter 2.

Abusing notation slightly, we identify $L^2([-\lambda,\lambda])$ with
a subspace of $L^2(\bR)$ via extension by $0$. Denote by $\lF$
restriction of $F$ to $\bR^+ \times L^2([-\lambda,\lambda])$ defined by equation
\ref{eqn:mod}. Then the FWI problem with support constraint,
described in the last paragraph, can be crudely phrased: given $d \in
L^2([t_{\rm min},t_{\rm max}])$, find $(m,w)$ in the domain of $\lF$ so that $\lF[m,w]
\approx d$.

Despite its simplicity, this problem exhibits the cycle-skipping
pathology characteristic of field-scale FWI. Specifically, we show
that the reduced mean square error function, the objective function of
FWI for this problem,
\begin{equation}
  \label{eqn:redms}
  \lerr[m,w;d]=\frac{1}{2}\|\lF[m]w-d\|^2/\|d\|^2,
\end{equation}
has local minimizers far from any global minimizer, even for noise-free
data (Theorem \ref{thm:fwi}).

This fact implies that local optimization must fail to solve a precise
version of the FWI problem:
\begin{quote} given $\epsilon, \lambda > 0$, fine $(m,w)$ in the
domain of $\lF$ so that
\begin{equation}
  \label{eqn:probstat}
  \|\lF[m]w-d\| \le \epsilon\|d\|.
\end{equation}
\end{quote}
In fact, we show
that``large residual'' local minima of $\lerr[\cdot,\cdot,d]$ exist,
at which its value is much greater than its global minimum
value. Therefore if $1/2\epsilon^2$ is near the global minimum value
of $\lerr[\cdot,cdot;d]$, then
local minimization of $\lerr[\cdot,\cdot;d]$ can fail to find a
solution of the inverse problem. This phenomenon is what is meant by
``cycle-skipping''.

Extended inversion involves an extension of the model-to-data map, in
this case $\lF$.  The extension used here consists in dropping the
support constraint on $w$, so the extended map is simply $F$ as
defined above ($\lF$ is a restriction of $F$, so $F$ is an extension
of $\lF$).

To compensate for the resulting indeterminacy in the
estimation of $m$, we add a penalty term to the objective, defined by
a penalty operator $A$ with domain and range $L^2(\bR)$: for $m \in
\bR^+,w\in L^2(\bR), d \in L^2([t_{\rm min},t_{\rm max}])$, define
\begin{eqnarray}
  r[m,w;d] & = & \|F[m]w-d\|/\|d\|, \label{eqn:rmserr}\\
  e[m,w;d] & = & \frac{1}{2}r[m,w;d]^2, \label{eqn:mserr}\\
  g[w] & = & \frac{1}{2}\|Aw\|^2/\|d\|^2, \label{eqn:pen} \\
  \Ja[m, w;d] &=& e[m,w;d] + \alpha^2 g[w] \nonumber\\
  &=& \frac{1}{2}(\|F[m]w-d\|^2 + \alpha^2 \|Aw\|^2)/\|d\|^2.
\label{eqn:penerr}
\end{eqnarray}

The normalizations by the data norm $\|d\|$ in the definitions
\ref{eqn:rmserr}, \ref{eqn:mserr}, \ref{eqn:pen}, and \ref{eqn:penerr} make the values
of $r$, $e$, $\alpha^2 g$, and $\Ja$ nondimensional, and simplifies
both the statements of theoretical results and the interpretation of
numerical examples.

Apart from dimensional correctness, the only obvious requirement on the penalty operator $A$ is that it
penalize 
nonzero values of $w$ for large $|t|$, so that minimizers $(m,w)$ of
$\Ja$ should have $w$
concentrated near $t=0$. Thus the penalty is a soft stand-in for the
support constraint. We
show that a good choice for $A$ is given (essentially) by
$(Aw)(t)=tw(t)$ (Proposition \ref{thm:rampgood}), and the
remainder of the discussion in this section makes this assumption. The
penalty weight $\alpha$ must be chosen somehow; we review a simple
algorithm for controlling $\alpha$ later in the paper.

Assume that $\mu>0$ and $d = F[m_*]w_* + n$, with $\mbox{supp }w_*
\subset [-\mu,\mu]$, and denote by $\eta = \|n\|/\|F[m_*]w_*\|$
the noise-to-signal ratio. We show that any local minimizer $(m,w)$
of $\Ja$ satisfies 
\begin{itemize}
\item If $\eta \le 0.6$,
  \begin{equation}
    \label{eqn:mbnd}
    |m-m_*| \le (1+f(\eta))\frac{\mu}{r},
  \end{equation}
  where $f(\eta)=2\eta+O(\eta^2)$ (Theorem \ref{thm:mnoiseres});
\item 
  $(m,{\bf 1}_{[-\lambda,\lambda]}w)$ is a solution of the FWI problem
  as stated above, that is,
  \begin{equation}
    \label{eqn:success}
    \lerr[m, {\bf 1}_{[-\lambda,\lambda]}w;d] \le \frac{1}{2}\epsilon^2,
  \end{equation}
  so long as
  \begin{eqnarray}
    \lambda & \ge & (2+f(\eta))\mu,\nonumber\\
    \epsilon &\ge & (4\pi r \alpha \lambda)^2 +\eta.
    \label{eqn:solcond}
  \end{eqnarray}
  (Theorem \ref{thm:ipnoisesuf})
\end{itemize}.
That is, the minimization of the extended objective $\Ja$ provides a
solution of the FWI problem, with constraints on the achievable
performance and bounds on the errors. All of these
bounds are shown to be sharp, and the constraints cannot be
significantly weakened. For example, in conclusion \ref{eqn:solcond}, the support radius $\lambda$ in
the projection onto $L^2([-\lambda,\lambda])$ must in general be at least twice the
support radius of the ``noise free'' wavelet $w_*$, plus an additional
allowance for noise.

Use of model extension to reveal the influence of earth mechanical
parameters on data is an old idea, and underlies several classes of
seismic data processing methods in widespread use for decades
\cite[]{geoprosp:2008}. Use of model extension to create optimization
principles for inversion is a more recent development
\cite[]{SymesCar:91,Plessix:00a,Symes:09,LuoSava:11,BiondiAlmomin:SEG12,LeeuwenHerrmannWRI:13,LiuSymesLi:14,Warner:14,ChaurisGP:14,Warner:16,LeeuwenHerrmann:16,Herve2017,HouSymes:Geo18}. Extended
inversion methods differ by the choice of additional degrees of
freedom, and by choice of penalty applied to eliminate them in the
final result. The approach studied in this paper belongs to the {\em
  source extension} variety: the extra parameters are provided by
permitting the energy source component of the experimental model to
have more, or less constrained, parameters than the experimental
design suggests. \cite{HuangNammourSymesDollizal:SEG19} give an
overview and taxonomy of source extension methods, with extensive
references.

Extended inversion is not the only alternative to straightforward
least-squares data fitting that may overcome cycle-skipping. For
example, evidence has recently emerged that the Wasserstein metric
from arising in the theory of optimal transport may provide a measure
of error between model-predicted and observed seismic data less
oscillatory than the mean-square
\cite[]{EngquistYang:GEO18,Metivier:GEO18}.

%In many cases, the constraint posed by membership of the second component $w$ in the subspace $\lW$ is an approximation to an idealized limit $\lambda \rightarrow 0$.
%It may be that $W_0 = \cap_{\lambda > 0}\lW$ is non-trivial, in which
%case $F$ may be viewed as an extension of $F_0 = F|_{ M \times
%  W_0}$. In that case, $F_0$ and $F$ realize the idealized extension
%structure discussed in many prior works on extended inversion (for
%example, \cite[]{geoprosp:2008}): $F_0$ is the physical modeling
%operator, and $W_0$ consists of second components of physical or
%feasible models. However, in many of these examples, $W_0$ is not
%computationally accessible, or - even if it is - gives results that
%closely approximate those obtained with $\lW$ for small $\lambda$. For
%other examples, for instance that  discussed later in this paper (and
%implicitly in \cite[]{Warner:16}, for instance), $W_0=\{0\}$ and the
%family $\{\lW: \lambda > 0\}$ actually embodies all of the necessary
%physics.

\section{Examples}
This section presents numerical examples of the estimates of $m$ and
$w$ obtained by minimization of $\Ja$. We have selected these examples
to illustrate the conclusions described in the Overview section.

While minimization of $\Ja$ might be tackled directly - by
alternating minimizations over $m$ and $w$, or by computing updates
for $m$ and $w$ simultaneously - such joint minimization performs
poorly, as \cite{YinHuang:16} has shown. The reason for this poor
performance is that $\Ja$ has dramatically different
sensitivity to $m$ versus $w$, as observed in the cited reference and elsewhere.
Instead, a nested approach, in which $w$ is
eliminated in an inner optimization,
generally gives far better numerical performance. This approach was
formalized by \cite{GolubPereyra:73,GolubPereyra:03} as the {\em
  Variable Projection Method} (``VPM''). Besides being the main
computational device used in our examples, it also plays a central
role in the theory developed in the next section.

The minimization of $\Ja$ over $w$ yields a function $\tJa$ of $m$
alone:
\begin{equation}
  \label{eqn:redexp0}
  \tJa[m;d] = \inf_w \Ja[m,w;d]
\end{equation}
Under appropriate conditions, met in our examples (Theorem
\ref{thm:norminv}, $\tJa$ has an alternate expression
\begin{equation}
  \label{eqn:redexp10}
  \tJa[m;d] = \Ja[m,\aw[m;d];d],
\end{equation}
in which $\aw[m;d] \in W$ is the solution of the normal
equation
\begin{equation}
  \label{eqn:norm0}
  (F[m]^TF[m]+\alpha^2A^TA)w= F[m]^Td.
\end{equation}

[ examples and discussion - Huiyi]

[ may need to repeat some of the explicit formulas for $\aw, e, g,
\tJa$ from the next section - refer to theorems]

%\section{Theory}

%\cite{FuSymes2017discrepancy} introduce a version of the discrepancy
%principle and its application to separable nonlinear least squares
%problems in penalty form \ref{eqn:basic}. The algorithm developed
%there is incomplete, however: it lacks  termination criteria. In this
%section, I review the discrepancy-based algorithm and introduce an
%appropriate stopping rule, thus completing the development of the
%algorithm presented in \cite{FuSymes2017discrepancy}.

\section{Findings}
This section describes the mathematical infrastructure necessary to
make precise the conclusions described in the Overview section, then
gives complete statements and proofs, along with those of
necessary auxiliary results.

We shall use the abbreviations ${\cal B}(X,Y)$ and ${\cal I}(X,Y)$ for 
the algebra of bounded linear operators from the Hilbert space $X$ to 
the Hilbert space $Y$, and its subalgebra of invertible
operators. All of the operators appearing in this discussion are
members of ${\cal B}(X,Y)$ for some choice of Hilbert spaces $X$ and
$Y$, and superscript $T$ denotes the transpose or adjoint of a such an
operator in the sense specified by its domain and range Hilbert structure.

\subsection{An Acoustic transmission Inverse Problem}
The data prediction or modeling operator for the single-trace acoustic
transmission problem has been defined above, in equation
\ref{eqn:mod}, which we restate here for convenience:
\begin{equation}
\label{eqn:mod1}
F[m]w(t)  = p(\bx_r,t) = \frac{1}{4\pi r}w\left(t-mr\right) 
\end{equation}
The slowness $m$ must be positive, as follows from basic acoustics,
and in fact reside in a range characteristic of the
material model: for crustal rock, a reasonable choice would be
$m_{\rm min}=0.125, m_{\rm max}=0.6$ s/km.

@@

\noindent{\bf Remark:} In computational practice, $W$ will have to be replaced by a
finite-dimensional subspace of $L^2(\bR)$. Many such choices will
implicitly limit the support of $w \in W$ to a bounded interval, say
$[T_{\rm min},T_{\rm max}]$. To maintain the surjective property,
these bounds should be chosen so that $[t_{\rm min}, t_{\rm max}]
\subset [T_{\rm min}+mr,T_{\rm max}+mr]$ for all $m \in M$, that is,
\begin{eqnarray}
  \label{eqn:dombounds}
  t_{\rm min} &\ge & T_{\rm min}+m_{\rm max}r,\nonumber\\
  t_{\rm max} &\le & T_{\rm max}+m_{\rm min}r.
\end{eqnarray}
We will ignore these computational necessities in this work,
maintaining the definition $W=L^2(\bR)$.

As mentioned earlier, $F[m]$ is surjective for every
$m \in M$. Since all possible data lie in the range of $F[m]$ for any
$m \in M$, some restriction of the domain
of $F$ is necessary in order that fitting the data constrain
$m$. The constraint proposed in the
Overview section involves a selection of a maxium support radius
$\lambda_{\rm max} >0$. Then define for $\lambda \in (0,\lambda_{\rm max}]$:
\begin{itemize}
\item $\lW = \{w \in W:
  \mbox{supp }w \subset [-\lambda,\lambda]\}$;
\item $\lF = F|_{M \times \lW}$.
\end{itemize}

In terms of this infrastructure, the inverse problem studied in 
this paper may be stated as a slightly refined version of the problem statement
\ref{eqn:probstat}:

>>>>

\title{Solution of an Acoustic Transmission Inverse Problem by
  Extended Inversion: Theory}
\author{William W. Symes\footnotemark[1]}
\address{\footnotemark[1]PO Box 43, Orcas, WA 98280 USA, {\tt symes@rice.edu}}

@@

\righthead{1D Acoustic Inversion: Theory}

\maketitle
\begin{abstract}
  A single-trace transmission inverse problem for the wave equation
  seeks to determine both the wave velocity in a homogenous acoustic
  medium and the transient waveform of an isotropic point source. The
  duration (support) of the source waveform and the source-to-receiver
  distance are assumed known. A least squares formulation of this
  problem exhibits the ``cycle-skipping'' behaviour observed in field
  scale problems of this type, with many local minima differing
  greatly from the global minimizer. This behaviour is eliminated by
  dropping the hard support constraint on the source waveform,
  replacing it by a soft penalty implemented as a weighted mean-square
  of the source waveform. For properly chosen weight function,
  penalizing nonzero values away from $t=0$, the velocity component of
  any stationary point differs from the global minimizer of the
  constrained least-squares formulation by a linear combination of the
  source waveform support radius and data noise-to-signal ratio. Given
  an estimate of data noise, the penalty weight can be dynamically
  adjusteded during iterative optimization to maximize predicted data
  accuracy and closely approximate a support-constrained source waveform.

@@

\section{Introduction}
Inverse problems based on wave equations (acoustic, elastic,
Maxwell's...) are commonplace in geophysics, nondestructive materials
testing, medical imaging, and other areas of science and engineering
in which wave motion plays an important role. Often these problems are
formulated as optimization of mean square (or other scalar measure)
functions of misfit between observed and predicted data. In many
contexts, such as exploration seismology, the computational size of
these problems is very large, and even the prediction (or modeling) of
the data requires use of high-performance computing
environments. Accordingly, rapidly convergent optimization algorithms
based on Newton's method and its relatives to update model parameters
are the numerical methods of choice. These however are local, in the
sense that they converge to stationary points of the objective
functions, usually near the initial model parameters. For
high-frequency (hence high resolution) data regimes, stagnation of
local optimization methods at physically irrelevant model vectors is
frequently observed. This ``cycle-skipping'' is one of the main
obstacles to widespread successful deployment of optimization-based
methods for inverse problems in wave propagation
\cite[]{GauTarVir:86,VirieuxOperto:09,Fichtner:10,Plessix:10,Schuster:17}.

This paper shows how a modification of least-squares data fitting via
{\em modeling operator extension} overcomes cycle-skipping in a simple
example. The problem studied here may be the simplest inverse wave
problem that exhibits cycle-skipping - in fact, essentially this
example is routinely used to illustrate cycle-skipping (see for
example \cite{VirieuxOperto:09}, Figure 7). Its data is a single trace
at a point in $\bR^3$ (``receiver location'') of the causal linear
acoustic pressure field, generated by an isotropic radiator acting at
another point (``source location'') at a known distance from the
receiver location. The acoustic material is supposed spatially
homogeneous, hence characterized by wave slowness (reciprocal
velocity) and density. The object of the inverse problem is to recover
the slowness and the time history of the isotropic radiator
(``wavelet'') from the recorded trace. In order that the data
constrain the slowness, the wavelet is presumed to have known compact
support. Thus the (``physical'') modeling operator has as its domain
the Cartesian product of an interval of positive slownesses, and the
subspace of $L^2(\bR)$ consisting of wavelets with the prescribed
support. Even for noise-free data (in the range of the modeling
operator), the obvious least squares problem formulated in terms of
this modeling operator exhibits many local minimizers having nothing
to do with the (``target'') slowness and wavelet used to create the
data - that is, cycle-skipping occurs (Theorem \ref{thm:fwi}).

Dropping the support constraint on the wavelet creates an extension of
the modeling map for which any data may be fit exactly, for any choice
of slowness. The link between data and slowness is restored by adding
a quadratic penalty on non-zero values of the wavelet outside of the
physical support interval to the extended mean-square data misfit. The
first main result of this paper is that for noise-free data, the
slowness component of {\em any} stationary point of this extended
penalty function differs from the target slowness by a multiple of the
target wavelet radius (Theorem \ref{thm:rampreallygood}). That is, cycle-skipping does not occur, and the
wavelet support radius determines the resolution of slowness. For
noisy data (that is, sum of a noise-free signal trace in the range of
the physical modeling operator and a square-integrable noise trace),
any stationary point differs from the target slowness by a combination
of the wavelet support radius and the noise-to-signal ratio (Theorem \ref{thm:mnoiseres}). Since the
wavelet component of the extended domain does not constrain support,
stationary points of the extended penalty function do not provide
solutions of the original, physical inverse problem, which includes
the support constraint. Such solutions can be obtained, with estimates
on support radius and data error, simply by truncating the wavelet
components of stationary points (Theorem \ref{thm:ipnoisesuf}).

The inverse problem studied here is far too simple to have any direct
use in applications. However it seems worthy of careful study for two
reasons. First, its simplicity allows an unusually complete account of
its properties. The present paper details its mathematical behaviour,
including explicit and sharp error estimates rarely available for nonlinear
inverse problems. A companion paper \cite[]{SymesChenMinkoff:21}
describes numerical examples that illustrate the mathematical
conclusions.

Second, the analysis reveals several important features shared with
similar approaches to other wave inverse problems, and so provides an
at least partial pattern for reaching similar conclusions with
immediate practical implications. For example, nested optimization
(variable projection method, \cite{GolubPereyra:03}) based on a model
decomposition into inner and outer variables seems to be very
important: in some cases, such as the simple problem presented here,
the decomposition is obvious (inner variable = wavelet, outer variable
= slowness), in others less so
\cite[]{geoprosp:2008,Terentyev:thesis}. The extended modeling
operator must be surjective, or at least have a dense range, {\em for
  each value of the outer variable}, so that data may be fit well even
for a poor initial guess of the outer variable. for the extended
inversion approach to be successful. The derivative of the extended
modeling operator with respect to the outer variable is
well-approximated by the composition of the extended modeling operator
itself and a pseudodifferential operator of order 1
\cite[]{Symes:IPTA14,tenKroode:IPTA14}. The choice of the penalty
operator, controlling the extended degrees of freedom, is critical: in
order to produce an objective immune from cycle-skipping, this penalty
operator must be (pseudo-)differential
\cite[]{StolkSymes:03}. Generally, the extended modeling operator is
not differentiable in the outer variable. However these last two facts
together imply the differentiability of the reduced objective function
from variable projection (Theorem \ref{thm:diffobj}, Appendix C). Because of the data-fitting
assumption, a straightforward algorithm for scaling the quadratic
penalty, based on the Discrepancy Principle, is available - see
\cite{FuSymes2017discrepancy,SymesChenMinkoff:21}, also Appendix
A. This scaling varies dynamically during iterative optimization, in
effect changing the objective function sporadically as the iteration
converges.

The approach explained here is an example of {\em extended source}
inversion, see \cite{HuangNammourSymesDollizal:SEG19} for a review,
and \cite{MetivierBrossier:SEG20} for a recent variation. Adding
degrees of freedom to the coefficients in the wave equation in various
ways produces so-called {\em extended model} inversion approaches,
which actually have a longer history in seismology - see
\cite{geoprosp:2008} for details. The need to overcome cycle-skipping
has spawned a great number of concepts, of which extension-based
inversion is only one. Least-squares inversion succeeds with
sufficiently good initial estimates of wave velocity, and the dominant
techniques in industrial and academic seismology exploit data other
than direct measurements of waves, for example measurement of wave
time-of-flight, to supply such initial estimates
\cite[]{VirieuxOperto:09,Fichtner:10,Schuster:17}. Use of a data
misfit measure other than least squares, such as versions of the
Wasserstein metric from optimal transport theory, also shows promise
\cite[]{Metivier:GEO18,EngquistYang:GEO18}.

The next section gives precise statements of the inverse problem
studied here, and its various components. The following sections
contain statements and proofs of the main results: the existence of
muliple stationary points of the least squares problem, properties of
the extended modeling operator and the reduced objective, choice of
the penalty operator, the main results mentioned above about
stationary points of the reduced objective, and the discrepancy-based
algorithm for adjusting the penalty scale. Three appendices justify
the discrepancy-based scale adjustment, review the link between
constrained support and spectrum provided by the Heisenberg inequality,
and expose the abstract structure underlying the differentiability of
the reduced objective.

I use the abbreviations ${\cal B}(X,Y)$ and ${\cal I}(X,Y)$ for 
the algebra of bounded linear operators from the Hilbert space $X$ to 
the Hilbert space $Y$, and its subalgebra of invertible
operators. All of the operators appearing in this discussion are
members of ${\cal B}(X,Y)$ for some choice of Hilbert spaces $X$ and
$Y$, and superscript $T$ denotes the transpose or adjoint of a such an
operator in the sense specified by its domain and range Hilbert structure.

\section{An Acoustic transmission Inverse Problem}
According to linear acoustics, the causal pressure field due to an
isotropic point radiator at $\bx_s \in \bR^3$ 
solves the the wave equation \cite[]{Frie:58}:
\begin{eqnarray}
  \label{eqn:awe}
  \left(m^2\frac{\partial^2 p}{\partial t^2} - \nabla^2\right) p(\bx,t) &=&
                                                                         w(t)\delta(\bx-\bx_s) \nonumber\\
  p(\bx,t)&=&0, t\ll 0.
\end{eqnarray}
The solution is well-known, see for instance
\cite{CourHil:62}, Chapter VI, section 12, equation 47:
\begin{equation}
  \label{eqn:homsol}
  p(\bx,t) = \frac{1}{4\pi |\bx-\bx_s|}w\left(t-m|\bx-\bx_s|\right),
\end{equation}
This trace of $p$ at $\bx_r \in \bR^3$ can be viewed as the result of
applying an m-dependent linear operator $F[m]$ to the wavelet $w$:
\begin{equation}
\label{eqn:mod}
F[m]w(t)  = p(\bx_r,t) = \frac{1}{4\pi r}w\left(t-mr\right) 
\end{equation}
The slowness $m$ must be positive, as follows from basic acoustics,
and in fact reside in a range characteristic of the
material model: for crustal rock, a reasonable choice would be
$m_{\rm min}=0.125, m_{\rm max}=0.6$ s/km.

@@

\noindent{\bf Remark:} In computational practice, $W$ will have to be replaced by a
finite-dimensional subspace of $L^2(\bR)$. Many such choices will
implicitly limit the support of $w \in W$ to a bounded interval, say
$[T_{\rm min},T_{\rm max}]$. To maintain the surjective property,
these bounds should be chosen so that $[t_{\rm min}, t_{\rm max}]
\subset [T_{\rm min}+mr,T_{\rm max}+mr]$ for all $m \in M$, that is,
\begin{eqnarray}
  \label{eqn:dombounds}
  t_{\rm min} &\ge & T_{\rm min}+m_{\rm max}r,\nonumber\\
  t_{\rm max} &\le & T_{\rm max}+m_{\rm min}r.
\end{eqnarray}
I will ignore these computational necessities in this work,
maintaining the definition $W=L^2(\bR)$.

As mentioned earlier, $F[m]$ is surjective for every $m \in M$. Since
all possible data lie in the range of $F[m]$ for any $m \in M$, some
restriction of the domain of $F$ is necessary in order that fitting
the data constrain $m$. The constraint employed is the specification
of a maxium support radius $\lambda_{\rm max} >0$. Then define for
$\lambda \in (0,\lambda_{\rm max}]$:
\begin{itemize}
\item $\lW = \{w \in W:
  \mbox{supp }w \subset [-\lambda,\lambda]\}$;
\item $\lF = F|_{M \times \lW}$.
\end{itemize}

In terms of this infrastructure, the inverse problem studied in 
this paper may be stated as

\begin{quote}
\noindent {\bf Inverse Problem:}
  given data $d \in D$, relative error level $\epsilon \in
  [0,1)$, and support radius $\lambda \in (0, \lambda_{\rm
    max}]$, find $(m,w) \in M \times \lW$ for which 
\begin{equation}
  \label{eqn:probstat0}  \|\lF[m]w-d\| \le \epsilon\|d\|,
\end{equation}
\end{quote}
